\MakeUppercase{Layered $H$-Partitions of $k$-Planar Graphs} \thanks{This work was partly funded by NSERC and MRI.}

### 2019-05-10 k-planar.tex
Getting closer....

A \emph{layering} of a graph $G$ is a partition $\mathcal{L}=\{V_0,V_1,\ldots,V_h\}$ of $V(G)$ so that there is no edge $vw\in E(G)$ such that $v\in V_i$, $w\in V_j$ and $|j-i|>1$.  For any partition $\mathcal{P}=\{S_1,\ldots,S_p$ of $V(G)\}$, a \emph{quotient graph} $H=G/\mathcal{P}$ has a $p$-element vertex set $V(H)=\{x_1,\ldots,x_p\}$ and $x_ix_j\in E(H)$ if and only there exists an edge $vw\in E(G)$ such that $v\in S_i$ and $w\in S_j$. To highlight the importance of the quotient graph $H$, we can $\mathcal{P}$ an $H$-partition and write this concisely as $\mathcal{P}=\{S_x : x\in V(H)\}$ so that each element of $\mathcal{P}$ is indexed by the vertex it creates in $H$.\footnote{An alternative definition of a layering of $G$ is a layered $H$-partition of $G$ where $H$ is a path.}  A layered $H$-partition $(\mathcal{L},\mathcal{P})$ of a graph $G$ consists of a layering $\mathcal{L}$ of $G$ and an $H$-partition $\mathcal{P}$ of $G$. The \emph{layered width} of $(\mathcal{H},\mathcal{P})$ is $\max\{|L\cap P|: L\in\mathcal{L},\, P\in\mathcal{P}\}$.

Layered $H$-partition of small layered width in which $H$ has some additional property are a useful tool for a number of graph problems. \citet{dujmovic.joret.ea:planar} show that, if each graph $G$ in some family $\mathcal{G}$ of graphs has a layered $H$-partition for which the layered-width of the partition and the treewidth of $H$ are each upper-bounded by some constant $c_\mathcal{G}$, then each of the following quantities are upper bounded by a constant, for every $G\in\mathcal{G}$: queue number \cite{X}, track number \cite{X}, etc \cite{X}, etc \cite{X}, etc \cite{X}. \citet{dujmovic.esperet.ea:planar} add non-repetitive chromatic number to this list.  \citet{dujmovic.joret.ea:planar} show that two common families of graphs have such $H$-partitions: planar graphs and (more generally) bounded-genus graphs.

@@

Under these definitions $0$-planar graphs are exactly planar graphs and $O$-plane graphs are plane graphs. Thus, like genus-$g$ graphs, $k$-planar graphs are a generalization of planar graphs (which are genus-$0$ graphs).  Unlike genus-$g$ graphs, however, the family of $k$-planar graphs is not minor closed. A graph $G'$ obtained from a $k$-planar graph $G$ by edge deletions and edge contractions may or may not be $k$-planar.

@@

Let $T$ be a rooted tree rooted at $r\in V(T)$. The \emph{depth} of a node $x\in V(T)$ is equal to the length of the path from $x$ to $r$ in $T$. A $P$ path in $T$ is \emph{vertical} if, for each integer $i$, $P$ contains at most one vertex of depth $i$.  The deepest vertex in a vertical path $P$ is called  $P$'s \emph{lower endpoint} and the other (shallowest) vertex in $P$ is called $P$'s \emph{upper endpoint}. For any node $v\in T$, we call each node on the path, in $T$, from $v$ to $r$ a \emph{$T$-ancestor} of $v$.

@@

The following property of a PS-Tree $K$ is an immediate consequence of \propref{children} that will be useful for us: \note{PM}{Do we still use this?}
\begin{compactenum}[(PR5)]
  \item \proplabel{ancestors} If $vw\in E(\Delta)$, $v\in V(N_x-F_x)$ and $w\in V(N_y-N_y)$ for some nodes $x,y\in V(K)$, then one of $x$ or $y$ is a $K$-ancestor of the other.
\end{compactenum}

@@

Let $G$ be a $k$-plane graph and let $G^+$ be the graph obtained as follows:
\begin{compactenum}
  \item Make $G$ planar by adding dummy vertices at each crossing to obtain a plane graph and then add edges to this graph to obtain a triangulation $G^+$.
  \item Add a complete graph $K_4$ that contains $G^{+}$ in one of its inner faces $F_0$ to obtain graph $G^{++}$.
  \item Add edges to complete $G^{++}$ to a triangulation $G^{+++}$.
\end{compactenum}
Let $r$ be the vertex of the $K_4$ not incident on $F_0$ and let $T$ be a BFS spanning tree of $G^+$ rooted at $r$. Observe that, for any edge $vw\in E(G)$, $\dist_{G^+}(v,w) \le k+1$ and therefore, for any two vertices $v,w\in V(G)$,
\begin{equation} 
  \dist_{G^+}(v,w) \le (k+1)\cdot \dist_G(v,w) \eqlabel{distance-preserving}
\end{equation}
For each integer $i\ge 0$, let $V_i=\{v\in V(G^+): \dist_{T}(r,v)=i\}$ and observe that $\mathcal{L}=V_0,V_1,\ldots,V_{h}$ is a layering of $G^+$ (but not of $G$).

Let $K=K(G^{+++},T,r)$ be a PS-tree and observe that the root $r_K$ of $K$ has $F_{r_K}=F_0$.  This implies:
\begin{compactenum}[(PR6)]
  \item \proplabel{all-in}  $V(G^+)\subseteq V(N_{r_K}-F_{r_K})$.
\end{compactenum}

For each vertex $v\in V(G^+)$, let $P_v=\{a\in V(K): v\in V(N_a-F_a)\}$.  
By (PR6), $P_{v}$ contains the root $r_K$ of $K$.  By \propref{children}, $K[P_v]$ is a vertical path (whose lower endpoint is the unique node $x\in V(K)$ such that $v\in Y_x$). Therefore $K[P_v]$ is a vertical path in $K$ that contains $r_K$.  We are about to repeatedly use the fact that the common intersection of two or more vertical paths is a vertical path.

@@

\begin{clm}
  For any $x\in V(H)$, and any integer $i\ge 0$, $|S_x\cap V_i|\le 24k+12$.
\end{clm}

@@

\begin{clm}
  The tree decomposition $\{B_x: x\in V(K)\}$ of $H$ contains no bag larger than $(3^{k+2}-3)/2$.
\end{clm}

@@

  \begin{figure}
    \begin{center}
        \includegraphics{figs/A-2}
    \end{center}
    \caption{The colour sequences defined by two minimal admissible $xa$-curves.}
    \figlabel{colouring}
  \end{figure}

@@

  By \clmref{bag-size}, $\mathcal{S}$ produces a graph $H=G/\mathcal{S}$ of treewidth $(3^{k+2}-1)/2$.  By \clmref{layered-width}, for each $x\in V(H)$ and each integer $i$, $|B_x\cap V_i|\le 24k+12$, so $|B_x\cap V_i'|\le (k+1)(24k + 12) = 24k^2+36k+12$. 
\end{proof}

>>>>

A \emph{layering} of a graph $G$ is a sequence $\mathcal{L}=\langle V_0,V_1,\ldots,V_h\rangle$ such that $\{V_0,V_1,\ldots,V_h\}$ is a partition of $V(G)$ and there is no edge $vw\in E(G)$ such that $v\in V_i$, $w\in V_j$ and $|j-i|>1$.  For any partition $\mathcal{P}=\{S_1,\ldots,S_p$ of $V(G)\}$, a \emph{quotient graph} $H=G/\mathcal{P}$ has a $p$-element vertex set $V(H)=\{x_1,\ldots,x_p\}$ and $x_ix_j\in E(H)$ if and only there exists an edge $vw\in E(G)$ such that $v\in S_i$ and $w\in S_j$. To highlight the importance of the quotient graph $H$, we call $\mathcal{P}$ an $H$-partition and write this concisely as $\mathcal{P}=\{S_x : x\in V(H)\}$ so that each element of $\mathcal{P}$ is indexed by the vertex it creates in $H$.\footnote{An alternative definition of a layering of $G$ is a layered $H$-partition of $G$ where $H$ is a path.}  A layered $H$-partition $(\mathcal{L},\mathcal{P})$ of a graph $G$ consists of a layering $\mathcal{L}$ of $G$ and an $H$-partition $\mathcal{P}$ of $G$. The \emph{layered width} of $(\mathcal{H},\mathcal{P})$ is $\max\{|L\cap P|: L\in\mathcal{L},\, P\in\mathcal{P}\}$.

Layered $H$-partitions of small layered width in which $H$ has some additional property are a useful tool for a number of graph problems. \citet{dujmovic.joret.ea:planar} show that, if each graph $G$ in some family $\mathcal{G}$ of graphs has a layered $H$-partition for which the layered-width of the partition and the treewidth of $H$ are each upper-bounded by some constant $c_\mathcal{G}$, then each of the following quantities are upper bounded by a constant, for every $G\in\mathcal{G}$: queue number \cite{X}, track number \cite{X}, etc \cite{X}, etc \cite{X}, etc \cite{X}. \citet{dujmovic.esperet.ea:planar} add non-repetitive chromatic number to this list.  \citet{dujmovic.joret.ea:planar} show that two common families of graphs have such $H$-partitions: planar graphs and (more generally) bounded-genus graphs.

@@

Under these definitions $0$-planar graphs are exactly planar graphs and $0$-plane graphs are plane graphs. Thus, like genus-$g$ graphs, $k$-planar graphs are a generalization of planar graphs (which are genus-$0$ graphs).  Unlike genus-$g$ graphs, however, the family of $k$-planar graphs is not minor closed. A graph $G'$ obtained from a $k$-planar graph $G$ by edge deletions and edge contractions may or may not be $k$-planar.

@@

Let $T$ be a rooted tree rooted at $r\in V(T)$. The \emph{$T$-depth} of a node $x\in V(T)$ is equal to the length of the path from $x$ to $r$ in $T$. A $P$ path in $T$ is \emph{vertical} if, for each integer $i$, $P$ contains at most one vertex of $T$-depth $i$.  The deepest vertex in a vertical path $P$ is called  $P$'s \emph{lower endpoint} and the other (shallowest) vertex in $P$ is called $P$'s \emph{upper endpoint}. For any node $v\in T$, we call each node on the path, in $T$, from $v$ to $r$ a \emph{$T$-ancestor} of $v$.

@@

% The following property of a PS-Tree $K$ is an immediate consequence of \propref{children} that will be useful for us: \note{PM}{Do we still use this?}
% \begin{compactenum}[(PR5)]
%   \item \proplabel{ancestors} If $vw\in E(\Delta)$, $v\in V(N_x-F_x)$ and $w\in V(N_y-N_y)$ for some nodes $x,y\in V(K)$, then one of $x$ or $y$ is a $K$-ancestor of the other.
% \end{compactenum}

@@

Let $G$ be a $k$-plane graph.  We will assume, for ease of exposition, that any point $p\in\R^2$ is involved in at most one crossing $(p,vw,xy)$ of $G$. This assumption is not critical since it can be enforced by a slight deformation of the edges of $G$.

Let $G^+$ be the graph obtained as follows:
\begin{compactenum}
  \item Add a dummy vertex at each crossing in $G$ to obtain a plane graph and then add edges to this graph to obtain a triangulation $G^+$.
  \item Add a complete graph $K_4$ that contains $G^{+}$ in one of its inner faces, $F_0$, and then add edges to the resulting graph to obtain a triangulation $G^{++}$.
\end{compactenum}
Let $r$ be the vertex of the $K_4$ not incident on $F_0$ and let $T$ be a BFS spanning tree of $G^+$ rooted at $r$. Observe that, for any edge $vw\in E(G)$, $\dist_{G^+}(v,w) \le k+1$ and therefore, for any two vertices $v,w\in V(G)$,
\begin{equation} 
  \dist_{G^+}(v,w) \le (k+1)\cdot \dist_G(v,w) \eqlabel{distance-preserving}
\end{equation}
For each integer $i\ge 0$, let $V_i=\{v\in V(G^+): \dist_{T}(r,v)=i\}$ and observe that $\{V_0,V_1,\ldots,V_{h}\}$ is a layering of $G^+$ (but not of $G$).

Let $K=K(G^{++},T,r)$ be a PS-tree and observe that the root $r_K$ of $K$ has $F_{r_K}=F_0$.  This implies:
\begin{compactenum}[(PR1)]\setcounter{enumi}{4}
  \item \proplabel{all-in}  $V(G^+)\subseteq V(N_{r_K}-F_{r_K})$.
\end{compactenum}

For each vertex $v\in V(G^+)$, let $P_v=\{a\in V(K): v\in V(N_a-F_a)\}$.  
By \propref{all-in}, $P_{v}$ contains the root $r_K$ of $K$.  By \propref{children}, $K[P_v]$ is a vertical path (whose lower endpoint is the unique node $x\in V(K)$ such that $v\in Y_x$). Therefore $K[P_v]$ is a vertical path in $K$ that contains $r_K$.  We are about to repeatedly use the fact that the common intersection of two or more vertical paths is a vertical path.

@@

\begin{clm}\clmlabel{width-of-g-plus}
  For any $x\in V(H)$, and any integer $i\ge 0$, $|S_x\cap V_i|\le 24k+12$.
\end{clm}

@@

\begin{clm}\clmlabel{bag-size}
  The tree decomposition $\{B_x: x\in V(K)\}$ of $H$ contains no bag larger than $(3^{k+2}-3)/2$.
\end{clm}

@@

  \begin{figure}
    \begin{center}
        \includegraphics{figs/A-2}
    \end{center}
    \caption{The colour sequences defined by two $i$-curves.}
    \figlabel{colouring}
  \end{figure}

@@

  By \clmref{bag-size}, $\mathcal{S}$ produces a graph $H=G/\mathcal{S}$ of treewidth $(3^{k+2}-1)/2$.  By \clmref{width-of-g-plus}, for each $x\in V(H)$ and each integer $i$, $|B_x\cap V_i|\le 24k+12$, so $|B_x\cap V_i'|\le (k+1)(24k + 12) = 24k^2+36k+12$. 
\end{proof}
